|

### 2021-01-14 article_format/extensions_format_revision.Rmd
Started major changes for resubmission

@@

Indeed, the magnitude of the bias in a multivariate model is somewhat more complex to derive, but it will be 
\begin{align}
\hat{\beta} & = \beta \frac{\Var(\theta)}{\Var(\theta) + \Var(\upsilon)}
\end{align}
where $\theta$ is just the residual of a regression in which the true underlying variable is regressed on all other covariates. In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ where $\tau$ is the intercept, and $\gamma_{1}$, $\gamma_{2}$ are the regression coefficients and $\theta_{t}$ is the residual [@Wooldridge2009, p. 318--320].

Normally it is not possible to reconstruct the bias since in cases where we have to rely on indicators, we would not know the true underlying variable. Furthermore, in the case of a fixed-effects model, the covariates are the unobserved time-invariant characteristics. However, because we are working with simulated data, we have everything we need. Going back to the results above, we can get the residuals of $x_{t}$ by either running a regression and saving the residuals, or we could skip a step and get them directly using the 'residual maker' matrix [@Ruettenauer2020] which is $\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}$ and $\bm{A} = \begin{pmatrix}\bm{\iota}_{n} & \bm{\alpha_{1}} & \bm{\alpha_{2}}\end{pmatrix}$ is the $n \times 3$ matrix of the covariates (plus a constant).

>>>>

# Model notation

The following is an outline of matrix notation for those that prefer it to the path diagrams and model syntax. There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
\begin{align}
\bm{y}^{*} & = \bm{\Lambda_{y}}^{*} \bm{\eta}^{*}, \\
\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*}, 
\end{align}
where $\bm{\eta}^{*} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{*} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{*} = (\bm{y}, \bm{x})^{\intercal}$. $\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables. $\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables. $\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables. Notice the $^{*}$ symbol is just meant to differentiate the vectors with them from those without them. That means, $\bm{\eta}^{*}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{*}$ holds the errors for the observed variables and the disturbances of the latent variables. $\bm{y}^{*}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{*}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{*}$. Lastly, $\bm{B}$ is a matrix that holds the regression coefficients.^[If we say that $p$ and $q$ stand for the number of observed dependent and independent variables, respectively, and $m$ and $n$ stand for the number of latent dependent and independent variables, respectively, then $\bm{\eta}^{*}$ and $\bm{\zeta}^{*}$ are $p + q + m + n$, $\bm{y}^{*}$ is $p + q$, $\bm{\Lambda_{y}}^{*}$ is $(p + q) \times (p + q + m + n)$ and $\bm{B}$ is $(p + q + m + n) \times (p + q + m + n)$ [@Bollen1989].]

This notation may be confusing at first, but it has advantages. First, it allows us the flexibility we need for the models. For example, it allows observed $\bm{x}$ to directly influence observed $\bm{y}$ (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators, see for example @Bollen1989, @Kline2016). It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly. These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables. It is also consistent with the notation used for these models in `lavaan`. In fact, `lavaan` switches automatically between matrix notations depending on the specified model. That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] It does have a potential disadvantage however. Besides being less intuitive than the typical $\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}$ notation, it means that by including the observed covariates in the stacked long vector $\bm{y}^{*}$, they are treated as another response variable with variances and covariances to be estimated by the model (instead of just using the sample statistics). This means that the assumption of multivariate normality (otherwise just imposed on the dependent variables) also applies to the independent ones [@Skrondal2004, p. 75]. This can be problematic for noncontinuous independent variables like sex, nationality dummies, marriage status (married/unmarried), etc. See @Skrondal2004 for more on this topic.

Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}). For that, we have the following matrix notation (with labels on the outside of the matrices):
\begin{align}
\begin{split}
\bm{y}^{*} & = \bm{\Lambda_{y}}^{*}\bm{\eta}^{*} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
 y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
 y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
 x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
 x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
 x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix},
\end{split} \label{eq:y}
\end{align}
\begin{align}
\begin{split}
\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} 
Notice in $\bm{\zeta}^{*}$, for the independent variables we could either write, for example $x_{t}$ or $\delta_{t}$. As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated. For the sake of simplicity, we will ignore this subtlety and refer to the observed variable from now on, keeping in mind that if the multivariate normality assumption holds, the estimated statistics will likely be sufficiently close to the sample ones for it to not make much of a difference.

Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet. We can remedy this by first putting the equation for $\bm{\eta}^{*}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.: 
\begin{align}
\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
\bm{\eta}^{*} - \bm{B}\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
(\bm{I} - \bm{B})\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
\bm{\eta}^{*} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*},
\end{align}
where $\bm{I}$ is the identity matrix. By substituting this back into the equation for the observed variables we get $\bm{y}^{*} = \bm{\Lambda_{y}}^{*}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*}]$, which works out to:
\begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align}
which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].

## Assumptions

What essentially differentiates an FE from an RE model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010]. The FE model assumes that $\E[\alpha x_{t}] \ne 0$. As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased. Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
\begin{align}
\bm{y}^{*}\bm{y}^{* \intercal} & = \E[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})^{\intercal}] \\
 & = \E[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\zeta}^{* \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal})] \\
 & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal} \\
 & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal}.
\end{align}
In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
\begin{align}
\bm{\Psi} & = \E
\begin{blockarray}{cccccccc}
 & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
 \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
 \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
 x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
 x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
 x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
 \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
 \end{block}
\end{blockarray}.
\end{align}
Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
\begin{align}
\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
 & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
 & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
 & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
\end{align}
This should make intuitive sense. From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual. For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})$. The rest of the model-implied covariance matrix results from $\bm{y}^{*}\bm{y}^{* \intercal}$.

@@

Indeed, the magnitude of the bias in a multivariate model is somewhat more complex to derive, but it will be 
\begin{align}
\hat{\beta} & = \beta \frac{\Var(\theta)}{\Var(\theta) + \Var(\upsilon)}
\end{align}
where $\theta$ is just the residual of a regression in which the underlying theoretical variable is regressed on all other covariates. In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ where $\tau$ is the intercept, and $\gamma_{1}$, $\gamma_{2}$ are the regression coefficients and $\theta_{t}$ is the residual [@Wooldridge2009, p. 318--320].

Normally it is not possible to reconstruct the bias since in cases where we have to rely on indicators, we would not have observed the underlying theoretical variable. Furthermore, in the case of a fixed-effects model, the covariates are the unobserved time-invariant characteristics. However, because we are working with simulated data, we have everything we need. Going back to the results above, we can get the residuals of $x_{t}$ by either running a regression and saving the residuals, or we could skip a step and get them directly using the 'residual maker' matrix [@Ruettenauer2020] which is $\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}$ and $\bm{A} = \begin{pmatrix}\bm{\iota}_{n} & \bm{\alpha_{1}} & \bm{\alpha_{2}}\end{pmatrix}$ is the $n \times 3$ matrix of the covariates (plus a constant).
